Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-08-05 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
Fix typo in leukemia experiment
Change color of add text to blue
A few more edits

In this work, we choose $L$ such that the domain $\Omega = \left[-L, L\right]$ contains all the inputs points $x_i$
\st{. Without loss of generality, we can assume all data points are contained in a symmetric interval around zero.}
{\color{blue}, and the set of input points $\{x_i\}_{i=1}^n$ has some kind of centering around zero.}
Let 
%
\begin{align}\label{eq_boundary_S}
S=\max_{i}|x_i|, 
\end{align}
%
then it follows that $x_i \in \left[-S, S\right]$ for all $i$. We now define $L$ as
%
\begin{align}\label{eq_boundary}
L=c \cdot S,
\end{align} 
%
where $S > 0$ represents  
%the half-range of the input space
%{\color{blue}the boundary of the zero-mean minimum symmetric domain $\Psi=[-S,S]$ containing the input values $\bm{x}$}
{\color{blue}the maximum absolute value of the input space}, 
and $c \geq 1$ is the proportional extension factor. In the following, we will refer to $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by 
%the half-range $S$ of the input space.
{\color{blue}the boundary $S$}. 
{\color{blue}Notice that $\bm{x}$ need not be symmetric around zero, but need to be contained in $\left[-S, S\right]$} (although it is recommended that $\bm{x}$ fits tightly the interval $\left[-S, S\right]$ to optimize computation and the iterative diagnostic presented in Section~\ref{subsec_diagnostics}).

@@

\item[A1.] Set the iteration index to $k=1$. Make an initial guess on the length-scale $\ell^{(k)}$. If there is no useful information available, we recommend to start with a large length-scale, {\color{blue}such as the normalized length-scale being about $\frac{\ell^{(k)}}{S} \in [0.5, 1]$}. 
%({\color{blue}Remember that $\ell$ here refers to the normalized length-scale $\frac{\ell}{S}$.})

@@

$2D$ and $3D$ synthetic functions were drawn from $2D$ and $3D$ GP priors, with input values $\bm{x}_i \in [-1,1]^2$ and $\bm{x}_i \in [-1,1]^3$, respectively, {\color{blue}and then it follows a input space boundary $S=1$ for both the $2D$ and $3D$ functions}. Squared exponential covariance functions with marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, $\ell_2=0.3$, and $\ell_3=0.4$, where $\ell_i$ is the length scale for the $i$'th dimension, were used in the GP priors. $200$ and $1000$ data points were sampled from the $2D$ and $3D$ drawn functions, respectively, and independent Gaussian noise with standard deviation $\sigma=0.2$ was added to the data points to form the final noisy sets of observations.

@@

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{align*}
p(y_i > t_i \mid f) &= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i \\
&=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{align*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$. Due to the predictor \textit{sex} ($x_2$) being a categorical variable (`1' for female and `2' for male), we apply indicator variable coding for the GP functions, in a similar way such coding is applied in linear models \citep{Gelman+Hill+Vehtari:2020:ROS}. The latent function $f(\bm{x})$, besides of being centered around a linear model, is composed of a general mean GP function, $h(\bm{x})$, defined for all observations, plus a second GP function, $g(\bm{x})$, that only applies to one of the predictor levels ('male' in this case) and is set to zero otherwise:
%
\begin{align*}
h(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_0) \right),  \\[1pt] 
%
g(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_1)\right),\\[1pt] 
%
f(\bm{x}) &= c + \bm{\beta} \bm{x} + h(\bm{x}) + \mathbb{I}\!\left[x_2=2\right]g(\bm{x}),
\end{align*}
%
where $\mathbb{I}\left[\cdot\right]$ is an indicator function. Above, $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model. $\theta_0$ contains the hyperparameters $\alpha_0$ and $\ell_0$ which are the marginal variance and length-scale of the general mean GP function, and $\theta_1$ contains the hyperparameters $\alpha_1$ and $\ell_1$ which are the marginal variance and length-scale, respectively, of a GP function specific to the male sex ($x_2=2$). Scalar length-scales, $\ell_0$ and $\ell_1$, are used in both multivariate covariance functions, assuming isotropic functions.
%
\begin{figure*}
\centering
\includegraphics[scale=0.70, trim = 0mm 5mm 0mm 0mm, clip]{fig15_posteriors_leukemia.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{fig15_posteriors_leukemia}
\end{figure*}

>>>>

In this work, we choose $L$ such that the domain $\Omega = \left[-L, L\right]$ contains all the inputs points $x_i$
\st{. Without loss of generality, we can assume all data points are contained in a symmetric interval around zero.}
{\color{blue}, and the set of input points $\{x_i\}_{i=1}^n$ is centered around zero.}
Let 
%
\begin{align}\label{eq_boundary_S}
S=\max_{i}|x_i|, 
\end{align}
%
then it follows that $x_i \in \left[-S, S\right]$ for all $i$. We now define $L$ as
%
\begin{align}\label{eq_boundary}
L=c \cdot S,
\end{align} 
%
where $S > 0$ represents  
%the half-range of the input space
%{\color{blue}the boundary of the zero-mean minimum symmetric domain $\Psi=[-S,S]$ containing the input values $\bm{x}$}
{\color{blue}the maximum absolute value of the input space}, 
and $c \geq 1$ is the proportional extension factor. In the following, we will refer to $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by 
%the half-range $S$ of the input space.
{\color{blue}the boundary $S$}. 
{\color{blue}Notice that $\bm{x}$ need not be symmetric around zero, but need to be contained in $\left[-S, S\right]$} (although it is recommended that $\bm{x}$ fits tightly the interval $\left[-S, S\right]$ to optimize computation and the iterative diagnostic presented in Section~\ref{subsec_diagnostics}).

@@

\item[A1.] Set the iteration index to $k=1$. Make an initial guess on the length-scale $\ell^{(k)}$. If there is no useful information available, we recommend to start with a large length-scale, {\color{blue}such as the normalized length-scale being in the interval $\frac{\ell^{(k)}}{S} \in [0.5, 1]$}. 
%({\color{blue}Remember that $\ell$ here refers to the normalized length-scale $\frac{\ell}{S}$.})

@@

$2D$ and $3D$ synthetic functions were drawn from $2D$ and $3D$ GP priors, with input values $\bm{x}_i \in [-1,1]^2$ and $\bm{x}_i \in [-1,1]^3$, respectively, {\color{blue}and the input space boundary is $S=1$ for both the $2D$ and $3D$ functions}. Squared exponential covariance functions with marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, $\ell_2=0.3$, and $\ell_3=0.4$, where $\ell_i$ is the length scale for the $i$'th dimension, were used in the GP priors. $200$ and $1000$ data points were sampled from the $2D$ and $3D$ drawn functions, respectively, and independent Gaussian noise with standard deviation $\sigma=0.2$ was added to the data points to form the final noisy sets of observations.

@@

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{align*}
p(y_i > t_i \mid f) &= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i \\
&=  1 - \Phi\! \left( \frac{\log(t_i)-f(\bm{x}_i)}{\sigma} \right),
\end{align*}
%
where $y_i>t_i$ denotes the unobserved survival time {\color{blue} and $\Phi$ is the standard normal cumulative distribution function}.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$. Due to the predictor \textit{sex} ($x_2$) being a categorical variable (`1' for female and `2' for male), we apply indicator variable coding for the GP functions, in a similar way such coding is applied in linear models \citep{Gelman+Hill+Vehtari:2020:ROS}. The latent function $f(\bm{x})$, besides of being centered around a linear model, is composed of a general mean GP function, $h(\bm{x})$, defined for all observations, plus a second GP function, $g(\bm{x})$, that only applies to one of the predictor levels ('male' in this case) and is set to zero otherwise:
%
\begin{align*}
h(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_0) \right),  \\[1pt] 
%
g(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_1)\right),\\[1pt] 
%
f(\bm{x}) &= c + \bm{\beta} \bm{x} + h(\bm{x}) + \mathbb{I}\!\left[x_2=2\right]g(\bm{x}),
\end{align*}
%
where $\mathbb{I}\left[\cdot\right]$ is an indicator function. Above, $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model. $\theta_0$ contains the hyperparameters $\alpha_0$ and $\ell_0$ which are the marginal variance and length-scale of the general mean GP function, and $\theta_1$ contains the hyperparameters $\alpha_1$ and $\ell_1$ which are the marginal variance and length-scale, respectively, of a GP function specific to the male sex ($x_2=2$). Scalar length-scales, $\ell_0$ and $\ell_1$, are used in both multivariate covariance functions, assuming isotropic functions.
%
\begin{figure*}
\centering
\includegraphics[scale=0.70, trim = 0mm 5mm 0mm 0mm, clip]{fig15_posteriors_leukemia.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{fig15_posteriors_leukemia}
\end{figure*}
